---
title: "00 - Per-sample loading, scDblFinder, SoupX, and QC"
author: "snRNA-seq publication pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
# Report runtime environment: locate the pipeline root and load the shared helper libraries.
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
# Locate the pipeline root from the Rmd file location or the working directory
# (compatible with both rendering and interactive runs; does not rely on .git/.Rproj).
input_dir <- tryCatch(
  dirname(normalizePath(knitr::current_input(dir = TRUE), mustWork = FALSE)),
  error = function(e) ""
)
root_candidates <- unique(normalizePath(
  c(input_dir, getwd(), file.path(getwd(), "snRNAseq_publication_pipeline")),
  mustWork = FALSE
))
pipeline_root <- root_candidates[
  file.exists(file.path(root_candidates, "R", "pipeline_helpers.R"))
][1]
if (is.na(pipeline_root)) {
  stop("Cannot locate the pipeline root containing R/pipeline_helpers.R", call. = FALSE)
}
knitr::opts_knit$set(root.dir = pipeline_root)
source(file.path(pipeline_root, "R", "pipeline_helpers.R"))
source(file.path(pipeline_root, "R", "step00_qc_helpers.R"))
```

## Pipeline overview

Each sample is processed independently: read the raw/filter matrices →
light pre-filtering (removes only empty droplets) →
scDblFinder doublet calling (run on the raw integer counts before SoupX; by default
doublets are only flagged, not removed) →
SoupX correction (reuses the scDblFinder clusters as soup clusters, so no extra
ad hoc clustering is needed) →
tissue-specific QC → merge by tissue and save.

All key parameters are provided by `config/pipeline_params.R`; heavy logic such as
input validation, per-sample processing, audit tables, and QC figure construction is
encapsulated in `R/step00_qc_helpers.R`. This script keeps only the workflow skeleton.

## Configuration

```{r configuration}
# Read key parameters from the centralized parameter file (seed, QC thresholds,
# doublet rate, SoupX settings, etc.).
source(file.path(pipeline_root, "config", "pipeline_params.R"))

# R packages required for this step; stop with an installation hint if any is missing.
required_packages <- c(
  "Seurat", "SeuratObject", "SoupX", "Matrix", "dplyr", "tidyr",
  "ggplot2", "future", "future.apply", "SingleCellExperiment",
  "SummarizedExperiment", "scDblFinder", "BiocParallel"
)
check_packages(required_packages)

# Limit BLAS/OpenMP to 1 thread to avoid stacking with the outer per-sample parallelism.
if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}
set.seed(seed)

# Parallelize across samples at the outer level; the backend is set by parallel_backend
# in config (default multicore/fork; on this WSL2 environment PSOCK hangs, so do not use
# multisession); scDblFinder stays serial inside each worker to avoid nested parallelism.
parallel <- setup_sample_parallel(n_workers, parallel_backend)

paths <- get_pipeline_paths(project_root = dirname(pipeline_root))
manifest <- make_sample_manifest()

# Optional per-sample override columns; NA = automatic/default behavior:
#   ExpectedDoubletRate in (0,1), e.g. 0.04 means 4%;
#   SoupXManualRho in [0,1], skips autoEstCont and uses this contamination rate directly.
manifest$ExpectedDoubletRate <- NA_real_
manifest$SoupXManualRho <- NA_real_
manifest$FilteredMatrixDir <- file.path(
  paths$data_dir, manifest$SampleID, manifest$SampleID, "outs", "filter_matrix"
)
manifest$RawMatrixDir <- file.path(
  paths$data_dir, manifest$SampleID, manifest$SampleID, "outs", "raw_matrix"
)
```

## Validate inputs and record run parameters

```{r validate_inputs}
# Hard-validate the experimental design (2 groups x 2 tissues x 3 replicates, both
# tissues per mouse) and input matrix integrity.
validate_manifest(manifest)

# Write the run status file (RUNNING); rewritten to SUCCESS at the end after a
# successful render.
run_started_at <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
run_status_file <- file.path(paths$table_dir, "00_RUN_STATUS.txt")
writeLines(
  c(
    "STATUS=RUNNING",
    paste0("STARTED_AT=", run_started_at),
    paste0("SCRIPT=", file.path(paths$pipeline_root, "00_data_loading_soupx_qc.Rmd"))
  ),
  run_status_file
)
write.csv(
  manifest,
  file.path(paths$table_dir, "00_sample_manifest.csv"),
  row.names = FALSE
)

# Record the key parameters of this run for auditing and reproducibility.
run_parameters <- data.frame(
  Parameter = c(
    "seed",
    "pre_qc_min_counts",
    "pre_qc_min_features",
    "doublet_rate_per_1000",
    "doublet_rate_sd",
    "doublet_rate_warning_difference",
    "doublet_rate_warning_fold",
    "remove_scDblFinder_doublets",
    "soupx_method",
    "soupx_round_to_integer",
    "parallel_backend",
    "parallel_workers"
  ),
  Value = c(
    seed,
    pre_qc_min_counts,
    pre_qc_min_features,
    doublet_rate_per_1000,
    doublet_rate_sd,
    doublet_rate_warning_difference,
    doublet_rate_warning_fold,
    remove_scDblFinder_doublets,
    "subtraction + autoEstCont",
    TRUE,
    parallel$backend,
    parallel$workers
  ),
  stringsAsFactors = FALSE
)
write.csv(
  run_parameters,
  file.path(paths$table_dir, "00_run_parameters.csv"),
  row.names = FALSE
)
run_parameters
```

## Run all samples

```{r run_all_samples}
# Derive a per-sample seed for reproducibility; restore the sequential plan in finally
# to avoid leaking parallel state.
sample_seeds <- seed + seq_len(nrow(manifest))
sample_results <- tryCatch(
  future.apply::future_lapply(
    seq_len(nrow(manifest)),
    function(i) {
      process_one_sample(
        sample_row = manifest[i, , drop = FALSE],
        sample_seed = sample_seeds[[i]],
        pre_min_counts = pre_qc_min_counts,
        pre_min_features = pre_qc_min_features,
        dbr_per_1000 = doublet_rate_per_1000,
        dbr_sd = doublet_rate_sd,
        dbr_warning_difference = doublet_rate_warning_difference,
        dbr_warning_fold = doublet_rate_warning_fold,
        remove_doublets = remove_scDblFinder_doublets,
        thresholds = qc_thresholds[[manifest$Tissue[[i]]]]
      )
    },
    future.seed = TRUE,
    # One sample per future (dynamic scheduling): the default chunking would statically
    # split the 12 samples into 6 workers x 2 samples with no load balancing, and an
    # error in one sample would also take down the other sample in the same chunk.
    future.chunk.size = 1
  ),
  finally = future::plan(future::sequential)
)
names(sample_results) <- manifest$SampleID

# Collect the four kinds of per-sample results: filtered object, sample-level summary,
# raw QC, and pre-final-filtering metadata.
seurat_list <- lapply(sample_results, function(x) x$obj)
sample_summary <- dplyr::bind_rows(
  lapply(sample_results, function(x) x$summary)
)
raw_qc_metadata <- dplyr::bind_rows(
  lapply(sample_results, function(x) x$raw_qc)
)
qc_metadata_before_final <- dplyr::bind_rows(
  lapply(sample_results, function(x) x$qc_before_final)
)
final_metadata <- dplyr::bind_rows(
  lapply(seurat_list, function(x) x[[]])
)

# Merge the raw and post-SoupX stage metadata into a per-cell QC audit table.
cell_qc_audit <- build_cell_qc_audit(
  raw_qc_metadata,
  qc_metadata_before_final,
  remove_scDblFinder_doublets
)

sample_summary
```

## Tables and QC audit trail

```{r write_tables}
# Write all QC tables: sample-level summary, per-cell audit (csv.gz),
# SoupX/doublet-specific tables, per-stage cell counts, QC thresholds, and
# failure-reason statistics.
write_qc_tables(sample_summary, cell_qc_audit, qc_thresholds, paths$table_dir)
```

## QC plots

```{r qc_plots, fig.width=14, fig.height=5}
# Build and display all QC figures: pre-/post-correction violin plots, UMI-vs-gene-count
# scatter, scDblFinder score distributions, and rho / doublet-rate bar plots.
qc_figs <- build_qc_figures(
  raw_qc_metadata,
  final_metadata,
  qc_metadata_before_final,
  sample_summary
)

print(qc_figs$raw_qc)
print(qc_figs$final_qc)
print(qc_figs$raw_scatter)
print(qc_figs$doublet_scores)
print(qc_figs$method_summary)

qc_figure_dir <- file.path(paths$fig_dir, "00_QC")
save_pub_plot(qc_figs$raw_qc, "00_qc_violin_before_filtering", qc_figure_dir, width = 14, height = 5)
save_pub_plot(qc_figs$final_qc, "00_qc_violin_after_filtering", qc_figure_dir, width = 14, height = 5)
save_pub_plot(qc_figs$raw_scatter, "00_raw_count_feature_scatter", qc_figure_dir, width = 14, height = 9)
save_pub_plot(qc_figs$doublet_scores, "00_scdblfinder_score_distributions", qc_figure_dir, width = 14, height = 9)
save_pub_plot(qc_figs$method_summary, "00_soupx_doublet_summary", qc_figure_dir, width = 12, height = 5)
```

## Merge and save one RDS per tissue

Cell IDs already carry their `SampleID` prefix, so the merge does not rely on
Seurat's automatic duplicate-name suffixes.

```{r merge_and_save}
# Merge sample objects by tissue and validate the merged result (metadata, cell IDs,
# sample set, total cell count).
spleen_samples <- manifest$SampleID[manifest$Tissue == "Spleen"]
bm_samples <- manifest$SampleID[manifest$Tissue == "BoneMarrow"]

sc_spleen <- merge_sample_objects(
  seurat_list[spleen_samples],
  project = "Spleen"
)
sc_bm <- merge_sample_objects(
  seurat_list[bm_samples],
  project = "BoneMarrow"
)

sc_spleen$Group <- normalize_group_levels(sc_spleen$Group)
sc_bm$Group <- normalize_group_levels(sc_bm$Group)

validate_merged_object(
  sc_spleen,
  expected_samples = spleen_samples,
  expected_cells = sum(
    sample_summary$CellsAfterFinalQC[sample_summary$Tissue == "Spleen"]
  ),
  tissue = "Spleen"
)
validate_merged_object(
  sc_bm,
  expected_samples = bm_samples,
  expected_cells = sum(
    sample_summary$CellsAfterFinalQC[
      sample_summary$Tissue == "BoneMarrow"
    ]
  ),
  tissue = "BoneMarrow"
)

saveRDS(
  sc_spleen,
  file.path(paths$rds_dir, "00_spleen_qc.rds")
)
saveRDS(
  sc_bm,
  file.path(paths$rds_dir, "00_bonemarrow_qc.rds")
)

tissue_summary <- data.frame(
  Tissue = c("Spleen", "BoneMarrow"),
  Samples = c(length(spleen_samples), length(bm_samples)),
  CellsAfterQC = c(ncol(sc_spleen), ncol(sc_bm)),
  stringsAsFactors = FALSE
)
write.csv(
  tissue_summary,
  file.path(paths$table_dir, "00_tissue_cell_counts_after_qc.csv"),
  row.names = FALSE
)

tissue_summary
```

## Reproducibility

```{r session_info}
# Record sessionInfo and rewrite the run status to SUCCESS.
session_info <- capture.output(sessionInfo())
writeLines(
  session_info,
  file.path(paths$table_dir, "00_session_info.txt")
)
writeLines(
  c(
    "STATUS=SUCCESS",
    paste0("STARTED_AT=", run_started_at),
    paste0("FINISHED_AT=", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
    paste0("SPLEEN_CELLS=", ncol(sc_spleen)),
    paste0("BONE_MARROW_CELLS=", ncol(sc_bm))
  ),
  run_status_file
)
sessionInfo()
```
